---
title: "South_america"
author: '239070104'
date: "2026-05-03"
output: html_document
---
## ELEVATION MAP OF SOUTH AMERICA IN R

This project generates a high-resolution elevation map of South America using a Digital Elevation Model (DEM) in R. The workflow integrates geospatial data processing and cartographic visualization techniques to produce a publication-ready terrain map.

The analysis begins by extracting South American country boundaries and automatically downloading elevation data from global DEM sources. The raster elevation data is then processed and transformed into a spatial format suitable for visualization. A hillshade layer is computed to enhance terrain relief and improve the three-dimensional perception of the landscape.

Finally, the elevation data is visualized using a hypsometric color palette combined with hillshade shading to create a realistic representation of topography. The map is refined with clean cartographic styling, including country borders, coordinate reference formatting, and a professional layout suitable for academic or professional presentation.

The final output is exported in high resolution, making it suitable for use in reports, publications, and professional platforms such as LinkedIn.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r} 
#Intalling packages 

library(terra)
library(elevatr)
library(sf)
library(ggplot2)
library(rnaturalearth)
```

```{r}
# ==============================
# 3. OBTENER SUDAMÉRICA1
# ==============================
world <- ne_countries(scale = "medium", returnclass = "sf")
south_america <- world[world$continent == "South America", ]

```

```{r}
#Lets plot the map
ggplot() +
  geom_sf(data = south_america, fill = "lightgreen", color = "black") +
  theme_minimal() +
  labs(title = "South America")
```
```{r} 
#South america in the world

plot(st_geometry(world), col = "lightgrey")
plot(st_geometry(south_america), col = "green", add = TRUE)
```
```{r}
# ==============================
# 4.DOWNLOADING DEM
# ==============================
dem <- get_elev_raster(
  locations = south_america,
  z = 5,   # resolution 
  clip = "locations"
)
```

```{r}
# ==============================
# 5. CONVERT TO TERRA
# ==============================
dem <- rast(dem)
```

```{r}
# ==============================
# 6. CREATE HILLSHADE
# ==============================
slope <- terrain(dem, "slope", unit = "radians")
aspect <- terrain(dem, "aspect", unit = "radians")

hillshade <- shade(slope, aspect, angle = 40, direction = 315)

```

```{r}
# ==============================
# 7. CONVERT TO DATAFRAME
# ==============================
dem_df <- as.data.frame(dem, xy = TRUE, na.rm = TRUE)
hill_df <- as.data.frame(hillshade, xy = TRUE, na.rm = TRUE)

colnames(dem_df)[3] <- "elevation"
colnames(hill_df)[3] <- "shade"
```

```{r}
#Cecking column names

names(dem_df)
names(hill_df)
```


```{r}
# ==============================
# 6. ADD COLOR PALETTE
# ==============================
hypso_colors <- c(
  "#0b3c5d",  # azul oscuro (muy bajo)
  "#3288bd",  # azul
  "#66c2a5",  # verde agua
  "#abdda4",  # verde claro
  "#e6f598",  # verde-amarillo
  "#fee08b",  # amarillo
  "#fdae61",  # naranja
  "#f46d43",  # rojo claro
  "#d73027",  # rojo
  "#a50026",  # rojo oscuro (alto)
  "#ffffff"   # blanco (picos altos)
)
```

```{r}
# ==============================
# 8. PLOTTING THE MAP
# ==============================
ggplot() +
  
  # Hillshade
  geom_tile(data = hill_df, aes(x = x, y = y, fill = shade), alpha = 0.35) +
  scale_fill_gradient(low = "black", high = "white", guide = "none") +
  
  # Elevation
  geom_tile(data = dem_df, aes(x = x, y = y, fill = elevation), alpha = 0.85) +
  scale_fill_gradientn(
    colours = hypso_colors,
    name = "Elevation (m)"
  ) +
  
  # Borders
  geom_sf(data = south_america, fill = NA, color = "black", linewidth = 0.2) +
  
  coord_sf(expand = FALSE) +
  
  labs(
    title = "Elevation Map of South America",
    subtitle = "Digital Elevation Model (DEM) with hillshade shading",
    caption = "Source: SRTM (elevatr) | Map by Nelson Arzu"
  ) +
  
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "aliceblue"),
    
    axis.title = element_blank(), #remove x y from axis 
    axis.text = element_text(size = 8), #coordinates font size
    axis.ticks = element_line(), #ticks active
    
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 12),
    legend.position = "right",
    legend.title = element_text(face = "bold")
  )
```


```{r}
#Saving the map in png format
ggsave(
  filename = "south_america_elevation_linkedin.png",
  plot = last_plot(),
  width = 8,      # vertical format
  height = 10,    # 4:5 ratio (LinkedIn)
  dpi = 400,      # high quality
  bg = "white"    # clean background
)
```

